---
title: "Cyclistic Bike-share Study Case"
author: "Lydia"
date: "2022-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How does a bicycle sharing business achieve success?
## What is Cyclistic Bike-Share?

In 2016, Cyclistic launched a successful bike sharing offering. Since then, the program has grown to a fleet of 5,824 bikes that are geo-tracked and locked into a network of 692 stations throughout Chicago. Bikes can be accessed from one station and returned to another at any time in the system.
Cyclistic differentiates itself from other companies by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive for people with disabilities and riders who cannot use standard two-wheeled bicycles.

## Stakeholders
1. Cyclistic users
2. Director of Marketing
3. Cyclistic Marketing Team
4. Cyclistic Executive Team

## Business Goals
### Design a marketing strategy aimed at converting regular riders into annual members.

## Table of Interest
### 1. Annual Members vs Casual Members
### 2. Cyclistic Annual Membership
### 3. Cyclistic Digital Media

## Data Cleaning
### Preparing Data
### Install Packages and Load Library

```{r install packages and load library}
install.packages("janitor")
install.packages("lubridate")
library(tidyverse)
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
```

### Load Dataset
We need to load all the dataset(The dataset is from January 2022-October 2022 --> the lastest dataset)

```{r Load dataset}
trip22_Jan <- read.csv("202201-divvy-tripdata.csv")
trip22_Feb <- read.csv("202202-divvy-tripdata.csv")
trip22_Mar <- read.csv("202203-divvy-tripdata.csv")
trip22_Apr <- read.csv("202204-divvy-tripdata.csv")
trip22_May <- read.csv("202205-divvy-tripdata.csv")
trip22_Jun <- read.csv("202206-divvy-tripdata.csv")
trip22_Jul <- read.csv("202207-divvy-tripdata.csv")
trip22_Aug <- read.csv("202208-divvy-tripdata.csv")
trip22_Sep <- read.csv("202209-divvy-publictripdata.csv")
trip22_Oct <- read.csv("202210-divvy-tripdata.csv")
```

### Combine all the Dataset to consolidate Analysis
Combine all that dataset and make new variable for the next futher analyze.

```{r Combine the dataset}
trip22_fill <- rbind(trip22_Jan, trip22_Feb, trip22_Mar, trip22_Apr, trip22_May, trip22_Jun, trip22_Jul, trip22_Aug, trip22_Sep, trip22_Oct)
```
### Check the dataset

```{r Check the dataset}
View(trip22_fill)
```

### Pick some dataset that we want to explore

```{r select the dataset}
trip22_fill2 <- trip22_fill %>% 
  select(ride_id, rideable_type, started_at, ended_at, start_station_name, end_station_name, member_casual)
```

### Review of the data and it's parameters

```{r review the dataset}
head(trip22_fill2, 10)
colnames(trip22_fill2)
nrow(trip22_fill2)
dim(trip22_fill2)
str(trip22_fill2)
summary(trip22_fill2)
```

### For started_at and ended_at must be created for Date and Time

```{r Convert the data type}
trip22_fill2$date <- as.Date(trip22_fill2$started_at)
trip22_fill2$month <- format(as.Date(trip22_fill2$date), "%m")
trip22_fill2$day <- format(as.Date(trip22_fill2$date), "%d")
trip22_fill2$year <- format(as.Date(trip22_fill2$date), "%Y")
trip22_fill2$day_of_week <- format(as.Date(trip22_fill2$date), "%A")
trip22_fill2$time <- format(trip22_fill2$started_at, format= "%H:%M")
trip22_fill2$time <- as.POSIXct(trip22_fill2$time, format= "%H:%M")
```

### create calculated field to isolate time spent on every ride.

```{r Created calculated field}
trip22_fill2$ride_length <- (as.double(difftime(trip22_fill2$ended_at, trip22_fill2$started_at))) /60
```

### Check data structure

```{r Check data structure}
str(trip22_fill2)
```

### Change data type for Ride Length

```{r Check data type--> ride length}
trip22_fill2$ride_length <- as.numeric(as.character(trip22_fill2$ride_length))
```

### Remove all blank entries for data set

```{r remove all blank entries}
trip22_fill3 <- trip22_fill2[!(trip22_fill2$start_station_name == "HQ QR" | trip22_fill2$ride_length<0),]
```

### Observe the new created column for the backup dataset

```{r observe}
summary(trip22_fill3$ride_length)
```

## Analyze Data
### Calculating the mean, median, max, min - figures to determine statistical speed of membership type

```{r calculating}
aggregate(trip22_fill3$ride_length ~ trip22_fill3$member_casual, FUN = mean)
aggregate(trip22_fill3$ride_length ~ trip22_fill3$member_casual, FUN = median)
aggregate(trip22_fill3$ride_length ~ trip22_fill3$member_casual, FUN = max)
aggregate(trip22_fill3$ride_length ~ trip22_fill3$member_casual, FUN = min)
```

### Order's days of week within new dataset for futher use

```{r orders days of week}
trip22_fill3$day_of_week <- ordered(trip22_fill3$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Create a Week Day

```{r create a week day}
trip22_fill3 %>% 
  mutate(day_of_week = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(num_of_rides = n())

View(trip22_fill3)
```

## DATA VISUALIZATION

```{r data visualization 1}
trip22_fill3$day_of_week  <- format(as.Date(trip22_fill3$date), "%A")
trip22_fill3 %>%                              #total rides broken down by weekday
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, day_of_week) %>%
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge") + 
  labs(x='Day of Week', y='Total Number of Rides', title='Rides per Day of Week', fill = 'Type of Membership') + 
  scale_y_continuous(breaks = c(250000, 400000, 550000), labels = c("250K", "400K", "550K"))
```
The rides per week show that the casual riders is peak on Saturday and Sunday, but the member riders is peak on Tuesday, Wednesday and Thursday

```{r data visualization 2}
trip22_fill3 %>%   #total rides broken down by month
  group_by(member_casual, month) %>%  
  summarise(total_rides = n(),`average_duration_(mins)` = mean(ride_length)) %>% 
  arrange(member_casual) %>% 
  ggplot(aes(x=month, y=total_rides, fill = member_casual)) + geom_col(position = "dodge") + 
  labs(x= "Month", y= "Total Number of Rides", title = "Rides per Month", fill = "Type of Membership") + 
  scale_y_continuous(breaks = c(100000, 200000, 300000, 400000), labels = c("100K", "200K", "300K", "400K")) + theme(axis.text.x = element_text(angle = 0))
```
The rides per month show that casual riders were a lot more active during the summer months than the long-term. 
Conversly, the winter months show very little activity on the part of the casual users. 
The long-term users are more active in the winter and spring months

```{r data visualization 3}
trip22_fill3 %>%    #looking at breakdown of bike types rented
  ggplot(aes(x = rideable_type, fill = member_casual)) + geom_bar(position = "dodge") + 
  labs(x= 'Type of Bike', y='Number of Rentals', title='Type of Bike Works The Most', fill = 'Type of Membership') +
  scale_y_continuous(breaks = c(500000, 1000000, 1500000), labels = c("500K", "1Mil", "1.5Mil"))
```
The breakdown of which type of bike is the most popular among either type of user. 
Showing among the two types of bikes classic and electric. 
Both types of memberships prefer using the classic bike more so than the electric bike. 
The long-term memebrs are also seen to be of the two types favours the classic bike.

```{r data visualization 4}
trip22_fill3 %>%        #Find the average time spent riding by each membership type per individul day
  mutate(day_of_week = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + labs(x='Days of the week', y='Average duration - Hrs', title='Average Ride Time per Week', fill='Type of Membership')
```
